Gradual Soundness: Lessons from Static Python

### 2021-12-20 pj-7.1/paper.tex
pj: sec 2.1 full draft

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
Relative to these, the \SP{} takes some interesting turns
because of the engineering context at Instagram.
The following contextual points are important to bear in mind as we present
notable aspects of the type system:
\begin{itemize}
  \item
    Performance is the bottom line.
    Unless \SP{} helps production code run faster, the team will be disbanded
    and reallocated.
    %% 2021-12-14: too extreme? and what about soundness?
  \item
    %% Q. why is the python so nominal? style guide? java background?
    The codebase makes heavy use of first-order classes and objects.
    First-class classes, objects, and even functions are rare.
  \item
    Instagram is / had been using Cython to improve the performance of select modules.
    Cython compiles Python source code to C extension modules for
    a faster Python [CITE].
    Instagram had converted several performance-critical modules to Cython prior
    to starting work on \SP{}.
    \SP{} replaced them.
    (How does this help understanding type system highlights?)
  \item
    The Instagram codebase is already typed.
    Developers use Pyre [CITE], an optional type checker for Python,
    to write type annotations and to drive IDE tools (e.g., autocompletion).
\end{itemize}

@@

The \SP{} solution is to provide two versions of every parameterized type,
both of which support efficient run-time checks.
One version is shallow.
The other is \emph{concrete}: complete, but new.
Consider \code{List[T]} as an example:
\begin{enumerate}
  \item
    \emph{Shallow type \code{List}}.
    The default type \code{List[T]} gets enforced with a simple check for lists that ignores
    the elements.
    Consequently, every static type like \code{List[int]} is merely a promise of \code{List}
    values at run-time.
  \item
    \emph{Concrete type \code{ChkList[T]}}.
    \SP{} introduces a new data structure and parameterized type \code{ChkList[T]}
    for concrete types.
    The data structure implements the list API and comes with a checked constructor that
    makes it impossible to create a \code{ChkList[int]} with non-integer elements;
    \cref{s:chkdict-impl} discusses the implementation further.
\end{enumerate}

@@

These options let programmers decide whether to change their code to benefit from
fully-sound types, or leave it as-is for a lower payoff.
Often, the shallow types are more attractive because concrete types impose
a ripple effect throughout the codebase.
If one type changes to concrete, then all data constructors that reach this type need
to change \emph{and then} all their clients need to use the concrete type as well.
\Cref{s:concrete-migration} discusses this migration cost of concrete types further.

\subsubsection{Mixed-Typed Class Hierarchies}

%% TODO very important, one of the main things IG does use!!!

The Instagram codebase uses lots of classes and objects.
Most of these are simple, Java-style: classes declared on
the top level and single inheritance.
It was therefore essential to get inheritance working across type boundaries.
Typed classes must be able to inherit from untyped classes,
and vice-versa untyped classes must be able to inherit from typed ones.

FILL literature? StrongScript Nom classic GT?

FILL either way, spell out what happens for these interactions.
Where are the checks needed, checks skipped.

\subsubsection{Types for Python}

Which types accommodate Python code? Which from PEP typing etc.

Unions are static only,
except for unions to false because those are common and can be implemented efficiently.

    Optional(T) is the only union. Others turn into Dynamic.

\subsubsection{Non-Highlights / Restrictions}

The \SP{} type system excludes include some features.
Each of the following is low priority for the team; that said, they are
ordered below rougly from most important (soonest to be implemented) to least.

\begin{itemize}
  \item \emph{Concrete Parameterized Classes}.
    Except for a handful of built-in types,
    parameterized classes are enforced only with shallow checks.
    In the near future, these classes will gain concrete versions
    for programmers to opt-in to.

  \item \emph{Callable Types, First-Class Classes}.
    The Instagram codebase rarely uses first-class functions, classes, or objects.
    Thus there is little need for either a \code{Callable} type to describe functions
    or first-class types for objects.
    When such values do appear, \SP{} assigns the dynamic type and spot-checks their
    uses.
    For example, every application of a first-class function gets wrapped in a check
    if the context expects a certain type.

  \item \emph{Method Overloading}.
    Methods cannot be overloaded.
    Each combination of a class and a name can refer to at most one method.
    This behavior comes from Python.

  \item \emph{Multiple Inheritance}.
    Multiple inheritance is not supported in a typed way.
    \Cref{s:impl} discusses this point further.

  \item \emph{Strong overrides}.
    Methods and fields may be overridden, but only with values of the same type.
    % In fact, \SP{} compiles fields using Python slots
    % declarations (\code{\_\_slots\_\_}).
    %
    % NOTE __ (dunder) methods may be overridden e.g. __getattribute__,
    % but the normal field access syntax skips any overrides (o.f)

  \item \emph{Recursive Types}.
    \SP{} does not support recursive types.
    Neither does Pyre or mypy.
    %% demand seems low ... baffling
    %% mypy = https://github.com/python/mypy/issues/731
    %% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/
\end{itemize}

@@

\subsubsection{Virtual Method Tables}

@@

\subsubsection{Unboxed Primitive}

>>>>

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
Relative to these, the \SP{} takes some interesting turns
because of the engineering context at Instagram.
The following contextual points are important to bear in mind as we present
notable aspects of the type system:
\begin{itemize}
  \item
    Performance is the bottom line.
    Unless \SP{} helps production code run faster, the team will be disbanded
    and reallocated.
    %% 2021-12-14: too extreme? and what about soundness?
  \item
    %% Q. why is the python so nominal? style guide? java background?
    The codebase makes heavy use of first-order classes and objects.
    First-class classes, objects, and even functions are rare.
  \item
    Prior to investing in \SP{}, Instagram used Cython to compile
    performance-critical code to C extension modules.
    Cython performed well, but mixing pure Python with a few compiled modules
    made for an awkward developer experience.
    \SP{} aims to match the Python workflow.
  \item
    The Instagram codebase is already typed.
    Developers use Pyre [CITE], an optional type checker for Python,
    to write type annotations and to drive IDE tools (e.g., autocompletion).
\end{itemize}

@@

The \SP{} solution is to provide two versions of every parameterized type,
both of which support efficient run-time checks.\footnote{Currently, only a handful
of built-in data structures provide shallow and concrete versions. User-defined
classes will soon follow this approach.}
One version is shallow.
The other is \emph{concrete}: complete, but new.
Consider \code{List[T]} as an example:
\begin{enumerate}
  \item
    \emph{Shallow type \code{List}}.
    The default type \code{List[T]} gets enforced with a simple check for lists that ignores
    the elements.
    Consequently, every static type like \code{List[int]} is merely a promise of \code{List}
    values at run-time.
  \item
    \emph{Concrete type \code{ChkList[T]}}.
    \SP{} introduces a new data structure and parameterized type \code{ChkList[T]}
    for concrete types.
    The data structure implements the list API and comes with a checked constructor that
    makes it impossible to create a \code{ChkList[int]} with non-integer elements;
    \cref{s:chkdict-impl} discusses the implementation further.
\end{enumerate}
%
These options let programmers decide whether to change their code to benefit from
fully-sound types, or leave it as-is for a lower payoff.

@@

Often, the shallow types are more attractive because concrete types impose
a ripple effect throughout the codebase.
If one type changes to concrete, then all data constructors that reach this type need
to change \emph{and then} all their clients need to use the concrete type as well.
\Cref{s:concrete-migration} discusses this migration cost of concrete types further.

\subsubsection{Gradual Class Hierarchies}
\label{s:inheritance}

One important feature of \SP{} is that class hierarchies can
mix typed and untyped code.
An untyped class may inherit from a typed one and vice-versa.
The point of this feature is to let programmers add types to a single class
independently of its ancestors and children.

% st-ecoop-2007 has no inheritance, object is collection of members simple (Abadi--Cardelli)

Gradual class hierarchies are rarely studied in the literature, especially
for a language where truly untyped classes can enter the mix.\footnote{Two
languages that provide separate hierarchies for untyped and (gradually) typed
classes are Thorn~\cite{wzlov-popl-2010} and SafeTS~\cite{rsfbv-popl-2015}.}
The closest related work, for Nom~\cite{mt-oopsla-2017}, is much more flexible.
\SP{} explores a simpler point in the design space, starting with
a few restrictions on types and inheritance:

\begin{enumerate}
  \item
    To benefit from types, classes must be declared on the module top level
    and may have at most one parent.
    Nested class declarations, first-class classes, and classes with
    multiple parents default to un-optimized Python behavior~(\cref{s:impl}).

  \item
    Methods cannot be overloaded.
    This restriction comes from Python.

  \item
    Methods and fields may be overridden in arbitrary ways by untyped code.
    In typed code, however, overrides must be compatible subtypes.
    For example, a method that returns a number cannot be overridden by a method
    that returns the dynamic type~(example: \cref{f:gg-failure-stat}).

    % In fact, \SP{} compiles fields using Python slots
    % declarations (\code{\_\_slots\_\_}).
    %
    % NOTE __ (dunder) methods may be overridden e.g. __getattribute__,
    % but the normal field access syntax skips any overrides (o.f)

\end{enumerate}
%
With this context in mind, every class member is either \emph{reliably} typed or untyped.
In particular, a typed member cannot be overridden by a less-precise one that uses
type dynamic.
\SP{} can therefore optimize dispatch from typed callers to typed methods
and use checked dispatch for other combinations.
One extra step arises when an untyped class overrides a typed method.
In this case, \SP{} creates a wrapper around the overriding method to check
that it computes type-correct results.
% Because these wrappers perform a simple first-order check, they
% are supported by an efficiently implementation that cooperates with
% \SP{}
The wrappers are handled efficiently by \SP{}'s vtables~(\cref{s:vtable}).

\subsubsection{Tailored Types}

\SP{} is selective about which types to support and which to omit.
In general, supported types are widely-used and come with an efficient
implementation.
Other widely-used types are deferred to Pyre for static checks until
the day when \SP{} adds support.\footnote{At present, the \SP{} team
has no plans to support all Pyre types. The top priority is to implement
core types efficiently.}

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Additional ``C types'' enable aggressive compilation of primitive
values~(\cref{s:ctypes}).
Union types are well-supported statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
but hardly supported at run-time.
Only binary unions of a type and false get enforced with a run-time check
(\code{Union[T, False]}, or \code{Optional[t0]}).

Three unsupported types bear special mention.
There are no first-class class or object types, as noted in \cref{s:inheritance}.
There are also no types for first-class functions (callables);
functions are common at Instagram, but first-class functions are rare.
Finally, \SP{} does not support recursive types due to low demand.
Neither Pyre nor mypy have recursive types either.
%% demand seems low ... baffling
%% mypy = https://github.com/python/mypy/issues/731
%% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/

@@

\subsubsection{Virtual Method Tables}
\label{s:vtable}

@@

\subsubsection{Unboxed Primitive}
\label{s:ctypes}
